Customizing Software by Direct Manipulation of Structured Data

### 2020-05-21 paper.tex
dm section

Table-driven customization embodies several design principles for
building customizable software systems, discussed in
Section~\ref{sec:themes}:

\begin{itemize}
\tightlist
\item
  \emph{Customization by direct manipulation}: End users should be able
  to customize an application by examining and modifying its data,
  rather than by writing imperative scripts.
\item
  \emph{Third-party semantic wrappers}: Typically, customization tools
  which don't rely on official extension APIs resort to offering
  low-level APIs for customization. Instead, we suggest a
  community-maintained library of semantic wrappers around existing
  applications, offering end users a more convenient semantic API for
  customization.
\item
  \emph{Gentle slope}: Customization tools should ideally offer a
  ``gentle slope'' from normal use to customization. We explore how our
  approach meets this goal by making some software customizations as
  easy as normal use of the software.
\end{itemize}

@@

Filling in dates for a flight search often requires a cumbersome
workflow: open up a separate calendar app, find the dates for the trip,
and then carefully copy them into the form. In Wildcard, the user can
avoid this by using a datepicker widget that shows the user's personal
calendar events{ (Figure~\ref{fig:expedia-demo}, Note B)}. The user can
directly click on the correct date, and it gets inserted into both the
spreadsheet and the original form.

@@

\begin{itemize}
\tightlist
\item
  reprise intro--explain direct manipulation
\item
  the key is that the DATA is being direct manipulated, not a script
\item
  it's essentially an alternate UI for an existing app, not a scripting
  env
\item
  argue for the benefit of multiple representations
  \citep{ainsworth1999}

  \begin{itemize}
  \tightlist
  \item
    abstraction
  \item
    different affordances
  \end{itemize}
\item
  address Tchernavskij's point about contrasting with instrumental
  interaction
\end{itemize}

\hypertarget{gentle-slope}{%
\subsection{Gentle slope}\label{gentle-slope}}

Term originally came from \citep{maclean1990}

\begin{itemize}
\tightlist
\item
  \citep{maclean1990} says ``some customizations should be as easy as
  using the app'', we have achieved that
\item
  here's where ``automation vs customization'' comes in: automation
  frames it as entering a whole new scripting environment; we frame it
  more as just an alternate UI that you can use.
\item
  backend APIs: terrible
\item
  the best would be right inline with the GUI (cite Scotty, instrumental
  work) but this has its own problems
\item
  we settle for a compromise: an alternate structured view. (explicitly
  contrast this clearly)
\item
  the CLI GUI thing
\end{itemize}

@@

\textbf{Third-party customization tools} enable customization without
using official extension APIs, enabling a broader range of
customizations on top of more applications. For example, web browser
extensions have demonstrated the utility of customizing websites through
manipulating the DOM, without needing explicit extension APIs to be
built in. However, third-party customization comes with a corresponding
cost: these tools can typically only operate at a low level of
abstraction, e.g.~manipulating user interface elements. This makes it
harder for end users to write scripts, and makes the resulting scripts
more brittle. (todo: support this claim more, provide examples)

\textbf{Semantic customization tools} use explicit extension APIs
provided by the application developer. Examples of this include
accessing a backend web API, or writing a customization in Applescript
that uses an application-specific API. The main benefit is that this
allows the extension author to work with meaningful concepts in the
application domain---``create a new calendar event'' rather than ``click
the button that says new event.''---which makes customizations easier to
build and more robust. However, this style limits the range of
extensions that can be built, to only those which the official plugin
API supports.

@@

One way to view this approach is as introducing a new abstraction
barrier into third-party extension. Typically, a third party
customization script combines two responsibilities: 1) mapping the
low-level details of a user interface to semantic constructs (e.g.,
using CSS selectors to find certain page elements), and 2) the actual
logic of the specific customization. (todo: could easily show examples
of this from browser extensions, Chickenfoot, etc) Even though the
mapping logic is often more generic than the specific customization, the
intertwining of these two responsibilities in a single script makes it
very difficult to share the mapping logic across scripts.

@@

Some of the challenges of writing a DOM scraping adapter are the same
ones involved with writing normal web scraping code, but the more
interactive nature of Wildcard introduces additional challenges. One
challenge is triggering updates to the spreadsheet data in response to
UI changes that happen after initial page load. Site adapters are
responsible for recognizing these changes by observing the DOM. So far,
we have been able to use event listeners and the MutationObserver API to
successfully observe changes, but it may prove challenging to observe
changes on some sites only through the DOM. Another challenge is
persisting updates to the DOM---some websites use virtual DOM frameworks
that can occasionally overwrite changes made by Wildcard. In practice,
we've managed to create an implementation that works around these issues
and results in a usable customization experience for all the websites
we've tried.

@@

This paper extends a workshop paper by Litt and Jackson \citep{litt2020}
which presented an earlier prototype of the Wildcard system. We build on
that work in this paper by rearchitecting the internal implementation of
the system around the table adapter abstraction, evaluating it on many
more websites, and by more fully characterizing the behavior and design
of the system.

@@

\begin{itemize}
\tightlist
\item
  customization tools

  \begin{itemize}
  \tightlist
  \item
    browser extensions
  \item
    ScrAPIr
  \item
    Thresher
  \item
    Sifter
  \item
    customization research tools: Chickenfoot, Coscripter
  \item
    Wildcard, like Chickenfoot, wants to hide HTML from users. But we
    show a structured data view, whereas Chickenfoot shows nothing
  \item
    visual customization tools:

>>>>

In Section~\ref{sec:themes}, we discuss some key themes from our work:

\begin{itemize}
\tightlist
\item
  \emph{Customization by direct manipulation}: End users should be able
  to customize an application by examining and modifying its data,
  rather than by writing imperative scripts.
\item
  \emph{Third-party semantic wrappers}: Typically, tools that don't rely
  on official extension APIs resort to offering low-level APIs for
  customization. Instead, we suggest a community-maintained library of
  semantic wrappers around existing applications, enabling end users to
  work with domain objects rather than low-level representations.
\end{itemize}

@@

Filling in dates for a flight search often requires a cumbersome
workflow: open up a separate calendar app, find the dates for the trip,
and then carefully copy them into the form. In Wildcard, the user can
avoid this by using a datepicker widget that shows the user's personal
calendar events{ (Figure~\ref{fig:expedia-demo}, Note B)}. The user can
directly click on the correct date, and it gets inserted into both the
spreadsheet and the original form.n

@@

\hypertarget{the-programming-cliff}{%
\subsubsection{The programming cliff}\label{the-programming-cliff}}

Hutchins, Hollan and Norman \citep{hutchins1985} describe a direct
manipulation interface as one that uses a model-world metaphor, rather
than a conversation metaphor. Rather than conversing with the system
about an assumed world, ``the world is explicitly represented'' and the
user can ``{[}act{]} upon the objects of the task domain themselves.''

Most GUIs today employ direct manipulation, but software customization
tools typically use an imperative programming model, which implements
the conversational metaphor rather than direct manipulation. For
example, in Applescript \citep{cook2007}, the scripting language for
customizing Mac OS applications, here is how a user retrieves a list of
of calendar names from the Calendar application:

\begin{verbatim}
tell application "Calendar"
  name of calendars
end tell
\end{verbatim}

Some customization environments (Mac Automator, Zapier) forego text
syntax and enable the user to connect programs and construct automations
by dragging and dropping commands. These environments still do not
constitute direct manipulation, though: the objects being manipulated
are in the domain of programming, not the domain of the task at hand.

It is a perfectly unreasonable choice to choose imperative programming
as the model for building customizations. Imperative programming
provides a high ceiling for possible customizations, and the model of a
sequence of commands is a natural fit for automation workflows which
simulate a series of steps taken by the user. However, there is a
serious drawback to this approach.

MacLean et al \citep{maclean1990} describe an ideal for user-tailorable
systems: a ``gentle slope'' from using to customizing, where small
incremental increases in skill lead to corresponding increments of
customization power. Requiring users to learn programming for
customization creates an abrupt ``cliff,'' where a significant learning
investment is required even to implement simple customizations. Another
goal of MacLean et al is that ``it should be as easy to change the
environment as it is to use it,'' at least for some subset of changes.
In scripting languages, even user-friendly ones like Applescript or
Chickenfoot, the experience of customization does not remotely resemble
the experience of use, so these systems can't meet this goal.

\hypertarget{a-gentler-slope}{%
\subsubsection{A gentler slope}\label{a-gentler-slope}}

With table-driven customization we aim to provide a gentler slope, by
using direct manipulation for software customization. The data shown in
the table view is the domain data from the original application. The
user makes changes to the data by selecting areas of interest in the
table, e.g.~sorting/filtering by clicking the relevant column header, or
adding annotation by clicking on the relevant row. These interactions
are common in GUI applications, and Wildcard therefore meets MacLean et
al's goal: some one-click customizations are as easy as using the
original application.

One aspect of directness which we have chosen not to maintain in
Wildcard is enabling customization in the context of the original user
interface, as explored by some other tools like Scotty
\citep{eagan2011}. We have found that augmenting the original UI with an
additional structured representation has numerous benefits. It provides
a consistent experience across all applications, and clearly shows the
user what structured data is available to work with.

Ainsworth et al's provide a helpful taxonomy of the value of multiple
representations \citep{ainsworth1999}. In their terms, Wildcard plays a
\emph{complementary role} by supporting a different set of tasks from
the original application, while displaying \emph{shared information}. We
also suspect that Wildcard helps \emph{constructs deeper understanding}
by \emph{subtraction}. By stripping away details and only showing the
essential data in an interface, Wildcard helps users think of new ways
of using that data, outside the specific constraints of the original
application.

It is important to note that there are many specific customizations that
can be achieved in systems like Chickenfoot \citep{bolin2005} that
cannot be reproduced in Wildcard. We consider this an acceptable
tradeoff in exchange for achieving a gentler slope of customization, and
we show in Section~\ref{sec:examples} and Section~\ref{sec:evaluation}
that our model can still implement many useful customizations in
practice. Also, there is sometimes a way to reframe an imperative script
in terms of our direct manipulation model; for example, a script that
iterates through rows in a page adding some additional information could
be reproduced using a formula in Wildcard.

@@

\emph{Third-party customization tools} enable customization without
using official extension APIs, enabling a broader range of
customizations on top of more applications. For example, web browser
extensions have demonstrated the utility of customizing websites through
manipulating the DOM, without needing explicit extension APIs to be
built in. However, third-party customization comes with a corresponding
cost: these tools can typically only operate at a low level of
abstraction, e.g.~manipulating user interface elements. This makes it
harder for end users to write scripts, and makes the resulting scripts
more brittle. (todo: support this claim more, provide examples)

\emph{Semantic customization tools} use explicit extension APIs provided
by the application developer. Examples of this include accessing a
backend web API, or writing a customization in Applescript that uses an
application-specific API. The main benefit is that this allows the
extension author to work with meaningful concepts in the application
domain---``create a new calendar event'' rather than ``click the button
that says new event.''---which makes customizations easier to build and
more robust. However, this style limits the range of extensions that can
be built, to only those which the official plugin API supports.

@@

One way to view this approach is as introducing a new abstraction
barrier into third-party extension. Typically, a third party
customization script combines two responsibilities: 1) mapping the
low-level details of a user interface to semantic constructs (e.g.,
using CSS selectors to find certain page elements), and 2) the actual
logic of the specific customization. Even though the mapping logic is
often more generic than the specific customization, the intertwining of
these two responsibilities in a single script makes it very difficult to
share the mapping logic across scripts.

@@

Some of the challenges of writing a DOM scraping adapter are the same
ones involved with writing normal web scraping code, but the more
interactive nature of Wildcard introduces additional challenges. One
challenge is triggering updates to the spreadsheet data in response to
UI changes that happen after initial page load. Site adapters are
responsible for recognizing these changes by observing the DOM. So far,
we have been able to use event listeners and the MutationObserver API to
successfully observe changes, but it may prove challenging to observe
changes on some sites only through the DOM. Another challenge is
persisting updates to the DOM---some websites use virtual DOM frameworks
that can occasionally overwrite changes made by Wildcard. Despite these
challenges, we've managed to create an implementation that works around
these issues in practice, and results in a usable customization
experience for all the websites we've tried so far.

@@

This paper extends a workshop paper by Litt and Jackson \citep{litt2020}
which presented an earlier prototype version of the Wildcard system. We
build on their work in this paper by rearchitecting the internal
implementation of the system around the table adapter abstraction,
evaluating it on many more websites, and by more fully characterizing
the behavior and design of the system.

@@

\begin{itemize}
\item
  Buttons \citep{maclean1990}: the original DM customization tool
\item
  ScrAPIr: direct manipulation for APIs!
\item
  customization tools

  \begin{itemize}
  \item
    browser extensions
  \item
    Thresher
  \item
    Sifter
  \item
    customization research tools: Chickenfoot, Coscripter
  \item
    Wildcard, like Chickenfoot, wants to hide HTML from users. But we
    show a structured data view, whereas Chickenfoot shows nothing
  \item
    visual customization tools:
